---
output:
  pdf_document: default
  html_document: default
---

# Using R with Single Variables

## Working with Diamonds Dataset
Loading the diamonds data set. 

I noticed that the diamonds data set will be listed as a 'Promise' in the workspace. This is a special object in R, and I need to run a command on the data to fully load the data set. I need to run summary in order to get the data appear in the environments.
```{r}
#must load the ggplot package first
library(ggplot2)

#loads the diamonds data set since it comes with the ggplot package 
data(diamonds)

summary(diamonds)
```

There are 53940 observations and 10 variables in the dataset. There are 3 ordered factors in the dataset: cut, color, clarity

```{r}
str(diamonds)
```
```{r}
levels(diamonds$color)
```
According to the documents, level 'D' is the best level for the diamonds.

### Price Histogram

Getting a histogram of the price shows a long-tailed distribution.
```{r}
qplot(data = diamonds, x = price,
      color = I('black'), fill = I('#099DD9'),
      binwidth = 100) +
  xlab('Price of Diamond') + 
  ylab('Number of diamonds in sample') +
  scale_x_log10() +
  scale_x_continuous(limits = c(0,15000), breaks = seq(0, 15000, 1000))
```

The shape of the price distribution is long-tailed.This means a large portion of the distribution is far from the 'head' or center of the distribution.

Running summary on price, we'll see:
```{r}
summary(diamonds$price)
```

Now, I want to get more statistics on the price of the diamonds. I answer questions like:

How many diamond cost less than $500?
To answer this question, first I get a summary of the price for those under $500. 
After executing the below command, the summary looks logical and gives true and false statements. True for those numbers below $500 and False for those above $500. 
```{r}
summary(diamonds$price < 500)

```

To make the above logical statement look nicer, I create a new variable called 'less_than_500' and assign it to 'NA'. 
Then I create a conditional statement: If the price is under $500, return 1; else return 0.
```{r}
less_than_500 <- NA
diamonds$less_than_500 <- ifelse(diamonds$price < 500, 1, 0)
diamonds$less_than_500 <- factor(diamonds$less_than_500)
summary(diamonds$less_than_500)
```

Next question, how many diamonds cost less than $250?
The same approach for the previous question will be applied here. 
```{r}
less_than_250 <- NA
diamonds$less_than_250 <- ifelse(diamonds$price < 250, 1, 0)
diamonds$less_than_250 <- factor(diamonds$less_than_250)
summary(diamonds$less_than_250)
```

The above calculations show that the conditional statement returns only those that do NOT cost less $250 dollars. 
This means there are no diamonds that cost less than $250.

Nest question, how many diamonds cost $15000 or more?
Let's apply the same approach in the 2 previous questions to this question as well.
```{r}
more_than_15000 <- NA
diamonds$more_than_15000 <- ifelse(diamonds$price >= 15000, 1, 0)
diamonds$more_than_15000 <- factor(diamonds$more_than_15000)
summary(diamonds$more_than_15000)
```

There are 1656 diamonds that cost $15000 or more.

### Cheaper Diamonds

Now, I want to explore the largest peak on the histogram I created earlier. I will use different binwidth, and limit the x-axis to observe the largest peak better.

According to my previous histogram, it's safe to say that the peak occured somewhere between 0-2000. I decided to set limits and breaks to get a closer look on the peak. I first started to set the limit(0,2000), and breaks=seq(0,2000,100), which gave me a much clear view that I can easily ignore anything under $300. 

Also, I could see there were no diamonds that cost $15000.

I decided to set my final limit between $300-1600.

```{r}
qplot(data = diamonds, x = price,
      color = I('black'), fill = I('#502FE4'),
      binwidth = 10) +
  xlab('Price of Diamond') + 
  ylab('Number of diamonds in sample') +
  scale_x_log10() +
  scale_x_continuous(limits = c(300,1600), breaks = seq(0, 1600, 100)) +
  ggsave('priceHistogram.png')
```

From this histogram, we can see the main peaks are between $600-900.

### Price by Cut Histograms

I want to break the histogram of diamonds prices by their cut. Let's have a look at the cut summary.
```{r}
summary(diamonds$cut)
```

There are 5 categories of cut: Fair, Good, Very Good, Premium, and Ideal. 

```{r}
qplot(data = diamonds, x = price, binwidth = 25) +
  xlab('Price Shown by Type of Cut') +
  ylab('Number of Diamonds in the Sample') +
  scale_x_continuous(limits = c(0, 15000), breaks = seq(0, 15000, 3000)) +
  facet_wrap(~cut)
```

Diving the prices based on the diamond cut, shows that the distribution looks similar in Ideal, Very Good and Premium categories; all long-tailed with a price peak. In the Good category, it is also long-tailed but the peak is not as much as in the first 3. In the Fair category the distribution looks uniform on the left with long tails on the right.

### Price by Cut

Now, let's answer some questions about the price of the diamonds based on their cuts:

Which cut has the highest priced diamond?

To answer this question, I use 'by' to get the summary of each cut. The information below shows that Very Good and Premium cuts have the maximum value of $18820.

```{r}
by(diamonds$price, diamonds$cut, summary)
```

Which cut has the lowest priced diamond?

Cuts Ideal and Premium have the lowest priced diamond $326.

Which cut has the lower median price?

Cut Ideal has the lowest Median price of $1810.

### Scales and Multiple Histograms

Getting a summary of the prices based on cuts shows that the distribution of the prices should be somewhat the same. However, in the histograms I created for prices based on cuts, we saw that only the distributions of Premium, Very Good and Ideal look the same. The distribution of Fair and Good look somehow uniform. 

Let's look into this more. I now create histograms of prices based on cuts, where the y-axis is not fixed. I do it by scaling the y-axis.

```{r}
qplot(x = price, data = diamonds,
      color = I('black'), fill = I('#25A00A')) + 
  xlab('Price Shown by Type of Cut') +
  ylab('Number of Diamonds in the Sample') +
  facet_wrap(~cut, scales="free_y")

```

Scaling the y-axis makes it different on some of the panels in the histogram. For example, for Fair cuts, the y-axis goes from 0 to over 200, whereas, in Premium it goes from 0 to about 3000.

### Price per Carat by Cut

I want to create a histogram of price per carat and facet it by cut. Since histograms only takes a single series and splits it into bins, for the 'x' variable, I use 'price/carat' to get the distribution for price per carat. I wrap it by cut, and add scales for the y-axis to get different values on the y-axis. I complete my histogram by scaling it using log10. After adding log10, I can see how the distribution has changed to a somehow normal one. 

```{r}
qplot(data = diamonds, x = price/carat,
      color = I('black'), fill = I('#F28D47')) +
  scale_x_log10() +
  facet_wrap(~cut, scales = "free_y")
```

### Price Box Plots

I want to investigate the price of diamonds using box plots, numerical summaries, and the following categorical variables: cut, clarity, or color.

#### Box Plot by Cut

To get the boxplot based on cut, I set the x-axis to show different cuts, and the y-axis to show the price of those cuts. 

```{r}
qplot(data = diamonds, geom = 'boxplot', 
      x = cut, y = price) +
  xlab('Diamond Cut') + 
  ylab('Price based on Cut')
  
```

These boxplots show there are a lot of outliers which fall over $10000 price. I need to get a closer look at the boxes without having to deal with the outliers. To do that, I use the 'coord_cartesian' which can magnify the boxes for me. 

The Cartesian coordinate system is the most familiar, and common, type of coordinate system. Setting limits on the coordinate system will zoom the plot (like you're looking at it with a magnifying glass), and will not change the underlying data like setting limits on a scale will.
Read more on: http://docs.ggplot2.org/0.9.3.1/coord_cartesian.html

Since the boxplots lie between 0 and $7000, I adjust my y-axis to limit the prices up to 7500. I also add a bit of color to the boxplots for a nicer look.

```{r}
qplot(data = diamonds, geom = 'boxplot', 
      x = cut, y = price,
      fill=cut) +
  xlab('Diamond Cut') + 
  ylab('Price based on Cut') +
  coord_cartesian(ylim = c(0, 7500))
```

The barplots show that the Premium cut has the largest range and highest median (i.e. the black horizontal line in the box), where the Ideal cut has the lower price.

#### Box Plot by Clarity

To get the boxplot based on clarity, I set the x-axis to show different clarities, and the y-axis to show the price of them.

I first generate my boxplot without using cartesian coordinates. This will give me the chance to see the ranges and where the outliers are. After investigating these things, I adjust my coordinates.

```{r}
qplot(data = diamonds, geom = 'boxplot', 
      x = clarity, y = price,
      fill=clarity) +
  xlab('Diamond Clarity') + 
  ylab('Price based on Clarity') +
  coord_cartesian(ylim = c(0, 6200))
  
```

The barplot show VS1 and VS2 have somehow the same range, with the Median of VS2 being slightly larger than VS1. The largest Median in price goes to SI2 clarity categorty.

#### Box Plot by Color

To get the boxplot based on clarity, I set the x-axis to show different colors, and the y-axis to show the price of them.

I first generate my boxplot without using cartesian coordinates. This will give me the chance to see the ranges and where the outliers are. After investigating these things, I adjust my coordinates.

```{r}
qplot(data = diamonds, geom = 'boxplot', 
      x = color, y = price,
      fill=color) +
  xlab('Diamond Color') + 
  ylab('Price based on Color') +
    coord_cartesian(ylim = c(0, 8000))

  
```

The barplots show that color E seems to have the lowest price, and the largest Median belongs to color J.

### Interquartile Range - IQR

Now I will work with the IQR() function and try to get some info regarding price and color of diamonds. IQR() computes the interquartile range of the x values. 

Let's first investigate a bit on the price range based on color. I use the 'by' command to get a summary of prices, categorized by color.

```{r}
by(diamonds$price,diamonds$color, summary)
```

Based on this summary, I can for example say that the first quartile(25%) of diamonds with color D has the price 911 dollors, and the third quartile(75%) of them are 4214 dollors.

#### What is the IQR for the diamonds with the worst color?

Based on the documentation for the diamonds dataset, in diamonds colors, 'J' is the worst and 'D' is the best.

```{r}
IQR(subset(diamonds, color=="J")$price)
```

#### What is the IQR for the diamonds with the best color?

```{r}
IQR(subset(diamonds, color=="D")$price)
```

### Price per Carat Box Plots by Color

I want to investigate the price per carat of diamonds across the different colors of diamonds using boxplots.

```{r}
qplot(data = diamonds, geom = 'boxplot', 
      x = color, y = price/carat) +
  xlab('Different Colors (From D to J)') +
  ylab('Price per Carat Based on Color') +
  coord_cartesian(ylim = c(2000, 6000))
    
```

Among different colors, color 'G' has the most range, and color 'I' has the lowest price. The highest Median goes to color 'H'.

### Carat Frequency Polygon

Using a frequency polygon, I want to find out what carat size has a count greater than 2000.

```{r}
qplot(data = diamonds, x = carat, binwidth = 0.1,
  geom = 'freqpoly') +
  scale_y_continuous(breaks = seq(0, 15000, 1000)) +
  scale_x_continuous(breaks = seq(0, 5, 0.2))

```

From this plot, I can see that for example 0.3 and 1.01 sizes (carats) have more than 2000 counts, while anything after 2.0 has fewer than 2000 counts.


